NumPy---Array Programming in Python

### 2020-01-25 body.tex
Restructure discussion

NumPy was initially developed by students, faculty, and researchers in their
spare time to provide a modern array programming library for Python.  They were
influenced by their experiences with powerful interactive programming languages
for scientific computing like Yorick \cite{munro1995using} as well as
commercial languages and environments like IDL and Matlab.  They also
frequently developed code from scratch to solve their own or their colleagues'
problems, often in low-level languages that preceded Python like Fortran
\cite{dongarra2008netlib} and C.

@@

This description of how APL users work easily applies to scientific Python
users \cite{kromberg2007apl}:
\begin{quote}
Users of APL tend to ``live in their data''. Interactive interpreters allow them
to inspect, almost feel their  way forward, discovering successful snippets
of code through experimentation and collecting them
into imperative or functional programs according to taste.
\end{quote}

NumPy spurred the development of the larger scientific Python ecosystem and
heralded the current era of wide-spread use of Python for scientific computing.
There were several factors that allowed this rapid growth and successful
development.

Python is already a good choice for standard programming tasks such as
cleaning data, interacting with web resources, and parsing text.
It has a wide range of libraries for different tasks.
Adding fast array operations and linear algebra allows the scientist to do all
their work with within the same language---and one that has the advantage of
being famously easy to learn and teach.

NumPy has strong culture of using software engineering practice to
improve collaboration and reduce error \cite{millman2014developing}.
The NumPy team was early in adopting distributed revision control and code
review to improve collaboration on code, and continuous testing that runs a
large battery of automated tests for every proposed change to NumPy.
The project has comprehensive, high-quality documentation, integrated with the
source code \cite{vanderwalt2008scipy,harrington2008scipy,harrington2009scipy}.

@@

Over time, and with the arrival of different technologies and array types, the
role of NumPy has changed from merely being an array provider, to being a
widely adopted array programming language.

>>>>

NumPy was initially developed by students, faculty, and researchers to
provide a modern array programming library for Python.  These
user-developers frequently had to write code from scratch to solve
their or their colleagues' problems---often in low-level languages
that preceded Python, like Fortran \cite{dongarra2008netlib} and C.
To them, the advantages of an interactive, high-level array library
were evident. Their design of this new tool was informed by their
experiences with powerful interactive programming languages for
scientific computing such as APL \cite{iverson1962programming} and
Yorick \cite{munro1995using}, as well as commercial languages and
environments like IDL and Matlab.

NumPy has subsequently seen prolific adoption, with millions of users,
first from research and later also industry.  It spurred the
development of the larger scientific Python ecosystem and heralded the
current era of wide-spread use of Python for scientific computing. The
reasons for its success are certainly complex, but we argue that they
are broadly of \emph{practical}, \emph{philosophical},
\emph{social}, and \emph{technical} origins.

@@

The \emph{practical} appeal of NumPy and its ecosystem is apparent
when you consider the following description of how APL users
work---one which
easily applies to scientific Python users too \cite{kromberg2007apl}:
\begin{quote}
Users of APL tend to ``live in their data''. Interactive interpreters allow them
to inspect, almost feel their  way forward, discovering successful snippets
of code through experimentation and collecting them
into imperative or functional programs according to taste.
\end{quote}

This modality of interactive scientific exploration would develop
further and eventually culminate in the Jupyter ecosystem of tools \fixme{citation needed}.

NumPy also benefited from being written in Python, a
general-purpose language well-suited to standard programming tasks such as
cleaning data, interacting with web resources, and parsing text.
Adding fast array operations and linear algebra allows scientist to do all
their work with within a single language---and one that has the advantage of
being famously easy to learn and teach.  In fact, Python's adoption as
a primary learning language in many universities \fixme{citation
needed} may well have further popularize NumPy as a tool of choice for
modern data science.

As an open source project, NumPy also did not cost money, and was
unencumbered by license servers, dongles, and the likes.

In terms of \emph{philosophical} appeal, NumPy is open source
software.  So, of course, are many other packages that are no longer
relevant, and by itself it does not motivate NumPy's sucess. However,
many developers would never have participated in building a closed
project.  Open source software also appealed the many proponents of
reproducible science, who wanted a transparent, introspectable stack
of software underlying their results.

\emph{Socially}, there was a sense of building something
impactful together, for the benefit of many others.  Participating in
such an endeavor, within a welcoming community of like-minded
indivuals, held a powerful attraction for many early contributors.

Throughout its development, NumPy has upheld strong \emph{technical}
principles.  It is a community of practice that enshrines a culture of
employing software engineering practice to improve collaboration and
reduce error \cite{millman2014developing}.  This culture is not only
adopted by leaders in the project, but also enthusiastically taught to
newcomers. The NumPy team was early in adopting distributed revision
control and code review to improve collaboration on code, and
continuous testing that runs a large battery of automated tests for
every proposed change to NumPy.  The project has comprehensive,
high-quality documentation, integrated with the source
code \cite{vanderwalt2008scipy,harrington2008scipy,harrington2009scipy}.

@@

Over time the role of the NumPy has changed. It is no longer merely an
array library, but also a standard API for tensor computation and a
central coordinating mechanism between array types and technologies.
NumPy has become a fully-fledged array programming paradigm, that
continues to drive numerical exploration and computation in scientific
Python.
